\mythesistitleNOTSYNPOSIS

### 2012-01-11 src/010chap1.tex
chapter 2 - started set system side of things. see diff for more details.
chapter 2 in edit still. see diff for changes.

As a special case, suppose all the apartments are on the same street
or if they are all lined up on a single path, the street map becomes a
tree that is just a path. Then the problem becomes what is called an
{\em interval assignment problem}. The idea of interval assignment may
not be obvious here; hence to see this, consider a different problem
in {\WSI} where the classes for these study groups courses need to be
scheduled during a day (or a week or any time period). Each study
group has a bunch of courses associated with it some of which may be
shared by two or more study groups. It is mandatory that a student who
is a member of a study group takes all the courses associated with
that group. There are slots during the day for classes to be held and
the problem is to allocate class slots to courses such that all the
classes of a study group are consecutive. It is debatable if this will
not hamper the attention span and memory retention rate of the
students but that is, regrettably, out of the scope of this
thesis. The parallels between this class allocation problem and the
accommodation problem can be seen as follows. The set $U$ here, are
the courses offered (say Course 101 {\coneohone}, Course 102
{\coneohtwo} and so on). In this variation of the problem, the
collection $\cF$ is the set of study groups but the study groups are
filled by course IDs (in place of students in the earlier
example). For instance, Course 101 is mandatory for all study groups
$\xLLL$, $\xGGG$, $\xBBB$, $\xTTT$ and Course 102 is mandatory for
only the $\xLLL$ group) and so on. The sequence of class slots for the
day (or week or any time period) is analogous to the street map in the
accommodation problem. It is quite obvious now why this version of the
problem (where the ``target graph'' is a path and not any
tree\tnote[minor]{Allowing any tree in this example could be seen as a
  scenario where there are parallel classes. A node falling in the
  path between two other nodes would mean that the corresponding is
  scheduled between the other two.}) is called an interval assignment
problem.

@@

\begin{definition}[\emph{Binary matrix}]
  Let $M$ be an $n \times m$ matrix. $m_{i,j}$ denotes its $(i,j)$th
  element, i.e. element at $i$th row and $j$th column. $M$ is a binary
  matrix if each of its element is 0 or \un.
  In other words, for all $i \in [n]$ and $j \in [m]$, $m_{i,j} \in
  \{0, \un\}$ 
\end{definition}

\begin{definition}[\emph{set system}]
\end{definition}

\begin{definition}[\emph{\Cop}]
  \tnote[important]{define cop here. refer to it in chap 2 or
    duplicate it using slash def }
  \begin{enumerate}
  \item \Cop
  \item Strong \cop
  \item \Cop order or \cop permutation
  \item \Cop for set systems\tnote{see mcc04}
  \end{enumerate}
\end{definition}

@@

\begin{definition}[\emph{\Crop}]
  \tnote[important]{define circ1 here. refer to it in chap 2 or
    duplicate it using slash def }
\end{definition}

\begin{definition}[\emph{Graph, Tree}]

\end{definition}

\begin{definition}[\emph{Maximal clique}]

\end{definition}

@@

The obvious first questions after being introduced to the consecutive
ones property of binary matrices are if COP can be detected
efficiently in a binary matrix and if so, can the COP permutation of
the matrix also be computed efficiently?  Recognition of COP in a
binary matrix is polynomial time solvable and the first such algorithm
was given by \cite{fg65}.  A landmark result came a few years later
when \cite{at72} discovered the families of forbidden submatrices that
prevent a matrix from having COP and most, if not all, results that
came later were based on this discovery which connected COP in binary
matrices to convex bipartite graphs. In fact, the forbidden
submatrices came as a corollary to the discovery that convex bipartite
graphs are AT-free on at least one of the partitions in
\cite{at72}\tnote{check the real meaning of AT FREE. }. The first linear
time algorithm for COP testing (COT) was invented by \cite{bl76} using
a data structure called PQ trees.  Since then several COT algorithms
have been invented -- some of which involved variations of PQ trees
\cite{mm96,wlh01,mcc04}, some involved set theory and ICPIA
\cite{wlh02,nsnrs09}, parallel COT algorithms\cite{as95,bs03,ly91} and
certifying algorithms\cite{mcc04}.

The construction of PQ trees in \cite{bl76} draws on the close
relationship of matrices with COP to interval graphs. A PQ tree of a
matrix is one that stores all row (column) permutations of the matrix
that give the COP orders (there could be multiple orders of rows or
columns) of the matrix. This is constructed using an elaborate linear
time procedure and is also a test for planarity\tnote{check check
  check. both interval graph and planarity in this paper?}.  PQR trees
is a generalized data structure based on PQ trees \cite{mm96,mpt98}.
\cite{tm05} describes an improved algorithm to build PQR
trees. \tnote{improv in terms of what?}\cite{wlh02} describes the
simpler algorithm for COT. Hsu also invented PC trees
\cite{wlh01}\tnote{This result first appeared inproc ISAAC92}
which is claimed to be much easier to implement. \cite{nsnrs09}
describes a characterization of consecutive-ones property solely based
on the cardinality properties of the set representations of the
columns (rows); every column (row) is equivalent to a set that has the
row (column) indices of the rows (columns) that have one entries in
this column (row). This is interesting and relevant, especially to
this thesis because it simplifies COT to a great degree. \tnote{it
  reduces the solution search space. fill in the blanks.}

@@

\tnote{ -- sect 4.1 in cite:d08phd has many results
  surveyed. hardness results, approx. results. results are usually for
  a class of matrices $(a,b)$ where number {\un}s in columns and rows
  are restriced to $a$ and $b$ . -- problem of flipping at most $k$
  entries of $M$ to make it attain COP. this is NP complete
  cite:b75-phd}\tnote{(1) scite:lb62 showed that interval graphs
  are AT-free.  describe AT (2) show the close relationship b/w COP
  and graphs sec 2.2, pg 31} \cite{at72} showed that a matrix that
does not have COP have certain substructures that prevent it from
having COP. Tucker classified these forbidden substructures into five
classes of submatrices. This result is presented in the context of
convex bipartite graphs which \cite{at72} proved to be
AT-free\tnote{ check this up. give details. - doms'}. By
definition, convex bipartite graph have half adjacency matrices that
have COP on either rows or columns (graph is biconvex if it has COP on
both)\cite{d08phd}. A half adjacency matrix is a binary matrix
representing a bipartite graph as follows. The set of rows and the set
of columns form the two partitions of the graph. Each row node is
adjacent to those nodes that represent the columns that have {\un}s in
the corresponding row. \cite{at72} proves that this bipartite graph
has no asteroidal triple if and only if the matrix has COP and goes on
to identify the forbidden substructures for these bipartite
graphs. The matrices corresponding to these substructures are the
forbidden submatrices.

Once a matrix has been detected to not have COP (using any of the COT
algorithms mentioned earlier), it is naturally of interest to find out
the smallest forbidden substructure (in terms of number of rows and/or
columns and/or number of entries that are {\un}s). \cite{d08phd}
discusses a couple of algorithms which are efficient if the number of
{\un}s in a row is small. This is of significance in the case of
sparse matrices where this number is much lesser than the number of
columns. $(*,\Delta)${\em -matrices} are matrices with no restriction
on number of {\un}s in any column but has at most $\Delta$ {\un}s in
any row. {\sc Min COS-R (Min COS-C), Max COS-R (Max COS-C)} are
similar problems which deals with inducing COP on a matrix. In {\sc
  Min COS-R (Min COS-C)} the question is to find the minimum number of
rows (columns) that must be deleted to result in a matrix with COP.
In the dual problem {\sc Max COS-R (Max COS-C)} the search is for the
maximum number of rows (columns) that induces a submatrix with
COP. Given a matrix $M$ with no COP, \cite{b75-phd} shows that finding
a submatrix $M'$ with all columns\tnote{check if b75 deals with
  COP col or COP row. also is it any submatrix with k less than r rows
  or submatrix must have all columns?} but a maximum cardinality
subset of rows such that $M'$ has COP is NP complete. \cite{hg02}
corrects an error of the abridged proof of this reduction as given in
\cite{gj79}.  \cite{d08phd} discusses all these problems in detail
giving an extensive survey of the previously existing results which
are almost exhaustively all approximation results and hardness
results. Taking this further, \cite{d08phd} presents new results in
the area of parameterized algorithms for this
problem\tnote{elaborate - what are the results?}.

>>>>

\subsection{Special case}
\label{sec:courseschedule}
As a special case of the study group accommodation problem, suppose
all the apartments are on the same street or if they are all lined up
on a single path, the street map becomes a tree that is just a
path. Then the problem becomes what is called an {\em interval
  assignment problem}. The idea of interval assignment may not be
obvious here; hence to see this, consider a different problem in
{\WSI} where the classes for these study groups courses need to be
scheduled during a day (or a week or any time period). Each study
group has a bunch of courses associated with it some of which may be
shared by two or more study groups. It is mandatory that a student who
is a member of a study group takes all the courses associated with
that group. There are slots during the day for classes to be held and
the problem is to allocate class slots to courses such that all the
classes of a study group are consecutive. It is debatable if this will
not hamper the attention span and memory retention rate of the
students but that is, regrettably, out of the scope of this
thesis. The parallels between this class allocation problem and the
accommodation problem can be seen as follows. The set $U$ here, are
the courses offered (say Course 101 {\coneohone}, Course 102
{\coneohtwo} and so on). In this variation of the problem, the
collection $\cF$ is the set of study groups but the study groups are
filled by course IDs (in place of students in the earlier
example). For instance, Course 101 is mandatory for all study groups
$\xLLL$, $\xGGG$, $\xBBB$, $\xTTT$ and Course 102 is mandatory for
only the $\xLLL$ group) and so on. The sequence of class slots for the
day (or week or any time period) is analogous to the street map in the
accommodation problem. It is quite obvious now why this version of the
problem (where the ``target graph'' is a path and not any
tree\tnote[minor]{Allowing any tree in this example could be seen as a
  scenario where there are parallel classes. A node falling in the
  path between two other nodes would mean that the corresponding is
  scheduled between the other two.}) is called an interval assignment
problem.

@@

\begin{definition}[\emph{Binary matrix}.]
  \label{def:binmatrix}
  Let $M$ be an $n \times m$ matrix. $m_{i,j}$ denotes its $(i,j)$th
  element, i.e. element at $i$th row and $j$th column. $M$ is a binary
  matrix if each of its element is 0 or \un.
  In other words, for all $i \in [n]$ and $j \in [m]$, $m_{i,j} \in
  \{0, \un\}$ 
\end{definition}

\begin{definition}[\emph{\Cop}]
    \label{def:copmatrix}
  \begin{enumerate}
  \item \Cop
  \item Strong \cop
  \item \Cop order or \cop permutation
  \item \Cop for set systems\tnote{see mcc04}
  \end{enumerate}
\end{definition}

@@

\begin{definition}[\emph{\Crop}]
  \label{def:cropmatrix}
  \tnote[important]{define circ1 here. refer to it in chap 2 or
    duplicate it using slash def }
\end{definition}

\begin{definition}[\emph{set system}]
    \label{def:setsystem}

    also include overlapping sets.
\end{definition}

\begin{definition}[\emph{Graph, Tree}]
  \label{def:graphtree}  
\end{definition}

\begin{definition}[\emph{Maximal clique}]
  \label{def:maxclique}    
\end{definition}

If $n$ is a positive integer, $[n]$ denotes the set $\{1,2,... n\}$.

@@

The obvious first questions after being introduced to the consecutive
ones property of binary matrices are if COP can be detected
efficiently in a binary matrix and if so, can the COP permutation of
the matrix also be computed efficiently?  Recognition of COP in a
binary matrix is polynomial time solvable and the first such algorithm
was given by \cite{fg65}.  A landmark result came a few years later
when \cite{at72} discovered the families of forbidden submatrices that
prevent a matrix from having COP and most, if not all, results that
came later were based on this discovery which connected COP in binary
matrices to convex bipartite graphs. In fact, the forbidden
submatrices came as a corollary to the discovery that convex bipartite
graphs are AT-free on at least one of the partitions in
\cite{at72}. The first linear time algorithm for COP testing (COT) was
invented by \cite{bl76} using a data structure called \PQtrees.  Since
then several COT algorithms have been invented -- some of which
involved variations of \PQtrees \cite{mm96,wlh01,mcc04}, some involved
set theory and ICPIA \cite{wlh02,nsnrs09}, parallel COT
algorithms\cite{as95,bs03,ly91} and certifying algorithms\cite{mcc04}.

The construction of \PQtrees in \cite{bl76} draws on the close
relationship of matrices with \COP to interval graphs. A PQ tree of a
matrix is one that stores all row (column) permutations of the matrix
that give the COP orders (there could be multiple orders of rows or
columns) of the matrix. This is constructed using an elaborate linear
time procedure and is also a test for planarity\tnote{check check
  check. both interval graph and planarity in this paper?}.  PQR trees
is a generalized data structure based on PQ trees \cite{mm96,mpt98}.
\cite{tm05} describes an improved algorithm to build PQR
trees. \tnote{improv in terms of what?}\cite{wlh02} describes the
simpler algorithm for COT. Hsu also invented PC trees
\cite{wlh01}\tnote{This result first appeared inproc ISAAC92}
which is claimed to be much easier to implement. \cite{nsnrs09}
describes a characterization of consecutive-ones property solely based
on the cardinality properties of the set representations of the
columns (rows); every column (row) is equivalent to a set that has the
row (column) indices of the rows (columns) that have one entries in
this column (row). This is interesting and relevant, especially to
this thesis because it simplifies COT to a great degree. \tnote{it
  reduces the solution search space. fill in the blanks.}

@@

\cite{at72} showed that a matrix that does not have COP have certain
substructures that prevent it from having COP. Tucker classified these
forbidden substructures into five classes of submatrices. This result
is presented in the context of convex bipartite graphs which
\cite{at72} proved to be AT-free in one of the partitions. By
definition, convex bipartite graph have half adjacency matrices that
have COP on either rows or columns (graph is biconvex if it has COP on
both)\cite{d08phd}. A half adjacency matrix is a binary matrix
representing a bipartite graph as follows. The set of rows and the set
of columns form the two partitions of the graph. Each row node is
adjacent to those nodes that represent the columns that have {\un}s in
the corresponding row. \cite{at72} proves that this bipartite graph
has no asteroidal triple in vertex partition corresponding to rows if
and only if the matrix has COP on columns and goes on to identify the
forbidden substructures for these bipartite graphs. The matrices
corresponding to these substructures are the forbidden submatrices.

Once a matrix has been detected to not have COP (using any of the COT
algorithms mentioned earlier), it is naturally of interest to find out
the smallest forbidden substructure (in terms of number of rows and/or
columns and/or number of entries that are {\un}s). \cite{d08phd}
discusses a couple of algorithms which are efficient if the number of
{\un}s in a row is small. This is of significance in the case of
sparse matrices where this number is much lesser than the number of
columns. $(*,\Delta)${\em -matrices} are matrices with no restriction
on number of {\un}s in any column but have at most $\Delta$ {\un}s in
any row. {\sc Min COS-R (Min COS-C), Max COS-R (Max COS-C)} are
similar problems which deals with inducing COP on a matrix. In {\sc
  Min COS-R (Min COS-C)} the question is to find the minimum number of
rows (columns) that must be deleted to result in a matrix with COP.
In the dual problem {\sc Max COS-R (Max COS-C)} the search is for the
maximum number of rows (columns) that induces a submatrix with
COP. Given a matrix $M$ with no COP, \cite{b75-phd} shows that finding
a submatrix $M'$ with all columns\tnote{check if b75 deals with
  COP col or COP row. also is it any submatrix with k less than r rows
  or submatrix must have all columns?} but a maximum cardinality
subset of rows such that $M'$ has COP is NP complete. \cite{hg02}
corrects an error of the abridged proof of this reduction as given in
\cite{gj79}.  \cite{d08phd} discusses all these problems in detail
giving an extensive survey of the previously existing results which
are almost exhaustively all approximation results and hardness
results. Taking this further, \cite{d08phd} presents new results in
the area of parameterized algorithms for this
problem\tnote{elaborate - what are the results?}.
